% Chapter 2 notes
% vim:fdm=marker

%{{{ ii-V space

\fntext{tf-1}{%
  \cite{aebersold:1974}. The Aebersold play-along series is a staple of jazz
  pedagogues; most contain a selection of tunes, along with a CD of a rhythm
  section so that students can practice with a recording. The \tf\ volume is
  number three of well over 100, and includes the phrase “the most important
  musical sequence in jazz!” on the cover.}

\fntext{tf-2}{%
  Despite our analysis in the previous chapter of “All the
  Things You Are” as taking place in a diatonic space, the ubiquity of \tfo\
  progressions means that many jazz musicians are apt to hear strings of \tf{}s,
  even when there is a clear diatonic pattern. Though some jazz musicians
  describe “Autumn Leaves” as a diatonic cycle, many may also describe it as a
  \tfo\ in \Bflat\ followed by a \tfo\ in the relative minor, connected by a
  major seventh chord.}

\fntext{tf-2a}{%
  The triangle on the C chord in this figure indicates a major seventh. We use
  the triangle rather than “\h{maj7}” or “\h{M7}” to save space and reduce
  clutter in the graphical representations.}

\fntext{tf-3}{%
  Lewin’s definition of a transformation network is in
  \headlesscite[196]{lewin:gmit}. For a relatively concise summary, see
  \cite[110--16]{rings:2011}.}

\fntext{tf-4}{%
  \headlesscite[Definition 1.3.1 (3)]{lewin:gmit}.}

\fntext{tf-5}{%
  \headlesscite[195--96 and throughout]{lewin:gmit}. See also
  \cite[6--8]{hook:2007}.}

\fntext{tf-6}{%
  Here and throughout this chapter, pitch classes are represented as mod-12
  integers, with C $= 0$; all calculations are performed mod-12.}

\fntext{tf-6a}{%
  This figure represents what Joseph Straus calls “transformational voice
  leadings” in his study of atonal voice leading; \headlesscite{straus:2003}.}

\fntext{tf-7}{%
  Hook himself makes this clear, noting that for any two sets S and
  T it is possible to define a single-type transformation in the set
  S $\cup$ T. He also notes that even when a single-type
  transformation is possible, ``the cross-type approach is often
  simpler and more natural,'' which certainly seems to be the case
  here. \headlesscite[5n8]{hook:2007}.}

\fntext{tf-8}{%
  It would be possible to form a \gis\ by considering all ordered triples as
  the group, as suggested above. While this is possible, defining an interval
  function in this group is much more difficult: such a function would need to
  account for the 36 ordered triples in \tf\ space (ii\tsup{7}, V\tsup{7}, and
  I\tsup{7} chords) as well as the many more (1692) that are not included in the
  space. Such a function is conceivable, but would not in any case reflect the
  musical realities \tf\ space is interested in portraying.}

\fntext{tf-9}{%
  \tf\ space is a directed graph, so in circumstances where the pitch-class
  distance metric is somehow not sufficient, we can instead rely on the
  standard way of measuring distance in a directed graph: by counting the
  number of edges in the shortest path between two chords. The distance from
  ii\tsup{7} of C to V\tsup{7} of \Eflat\ is then 4: ii to V in C (1 edge),
  then 3 $T_5$s to V of \Eflat.}


\fntext{tf-10}{%
  These changes are taken from \emph{The Real Book}, and reflect what is
  played on the album \emph{Cornbread}.\nocite{morgan:cornbread} In this figure,
  the circle indicates the tonic, while the numbers on the labels indicate the
  order of transformations.}

\fntext{tf-11}{%
  We will return to the concept of tritone substitutes in the next section.}

\fntext{tf-12}{%
  For a study of many different musical spaces (and a defense of their
  use), see \cite[Chapter 1 and throughout]{hook:mst}.}

\fntext{tf-13}{%
  Music analysis is in many ways similar to the field of data visualization:
  both often involve revealing the underlying structure of what might otherwise
  seem like an undifferentiated stream of data. As Edward Tufte tells us,
  visualizations are often ``more precise and revealing'' than other,
  mathematical means of analyzing data. \headlesscite[13--14]{tufte:2001}.}

%}}}

%{{{ tritone subs

\fntext{tft-1}{%
  The tritone substitution has been discussed extensively in the literature,
  so we will not discuss it at any length here. See, for example,
  \cite{biamonte:2008}; \cite{martin:1988}; and \cite{tymoczko:2011}.}

\fntext{tft-2a}{%
  A similar diagram can be found in \cite[5]{pohlert:bh}, and can be seen
  implicitly in Figure 1-1 of \cite{martin:1988}.}

% xxx citations here!
\fntext{tft-2}{%
  Möbius strips have a surprising number of applications in music theory. See,
  for example, % xxx citations needed!
}

\fntext{tft-2b}{%
  Determining the structure of the underlying transformation graph of this
  network is straightforward, so I have not included a figure of it here.}

\fntext{tft-3}{%
  Though we are defining chords as ordered triples in this chapter, I have
  included the fifth in this description to highlight the relationship to the
  triadic \textsc{slide}, which maintains the root and fifth of a triad while
  changing the quality of the third.}

% xxx put this section number in later!
\fntext{tft-3a}{%
  The \textsc{slide} transformation was introduced by David Lewin
  (\headlesscite[178]{lewin:gmit}), but has since become a part of of the
  standard set of Neo-Riemannian transformations. \slideS
  is defined here only as a transformation from I\tsup{7} chords to
  ii\tsup{7} chords, but of course the triadic \textsc{slide} is an involution
  (two successive applications of \textsc{slide} to any triad will result in the
  starting triad). We can define \slideS so that it acts in a similar way, but
  we will delay that possibility until Section \ref{}.}

% xxx look up this faure passage
\fntext{tft-4}{%
  The \slideS\ transformation can be found in a chromatic sequence in the
  second movement of the Fauré string quartet. (I am thankful to Julian Hook
  for pointing this example out to me.)}

\fntext{tft-5}{%
  % xxx citation of "Bird Blues"
  This progression is often known as the ``Bird Blues,'' though Mark Levine
  calls it the ``descending blues'' (\headlesscite[228]{levine:1995}). Like many
  sets of Parker changes, several different versions exist; the changes here
  represent a mediation of these sources. \emph{The Real Book} gives \h{Am7} (a
  vi chord) instead of \h{FM7} in m.~10; Levine's \emph{Jazz Theory Book} gives
  \h{Db7} (a tritone substitute) instead of \h{G7} in the second half of m.~3;
  and the \emph{Charlie Parker Omnibook} omits both \h{D7} chords (the first in
  the second half of m.~7, the second in m.~11) and the tonic in m.~11 is an
  F7.\nocite{aebersold:omnibook} Most of these differences are minor, and over
  the course of a recorded performance the changes might vary among all of these
  versions. Other compositions that contain this progression include Parker's
  own ``Confirmation,'' Sonny Stitt's ``Jack Spratt,'' and Toots Thielemans's
  ``Bluesette.''}

\fntext{tft-6}{%
  The major-minor seventh chord as a stable chord is characteristic of the
  blues; see, for example, \cite[158--59 and throughout]{mcgowan:2011}. This
  fact is somewhat obscured in \tf space, since major-minor sevenths appear
  in the space only as V\tsup{7} chords; we will return to this limitation
  later in this chapter.}

\fntext{tft-7}{%
  At least, impossible if we want to stay within the three sets we are
  studying in this chapter. Like the other transformations we have defined, EC
  is an admissible transformation on the set of all mod-12 ordered triples: if
  we begin with a \h{Dm7} chord, EC gives us $(2, 5, 0) \ECarrow\ (7, 10, 4)$,
  which of course is not a major, minor, or dominant seventh chord.}

\fntext{tft-8}{%
  A ``turnaround'' is what jazz musicians call a short progression that leads
  from a chord (often a tonic chord) back to itself. They appear most commonly
  at the ends of forms, and provide harmonic interest during solos, when a
  player might play several choruses in a row. The \tf appears frequently in
  this formal location, as do many of its variants: \mbox{vi--ii--V},
  \mbox{iii--VI--ii--V}, \mbox{iii--\h{bIII7}--ii--\h{bII7}}, etc.}

%}}}

%{{{ Extensions

\fntext{tfe-1}{%
  The extensions used in the dominant chord in a minor \tf is quite flexible:
  Aebersold's \emph{II--V\tsup{7}--I Progression} gives the quality as \h{7s9},
  but Mark Levine usually gives the chord symbol simply as ``alt.'' Levine
  includes the minor \tf in the category of ``melodic minor scale harmony,'' and
  ``alt.'' is short for the ``altered scale'' (the seventh mode of melodic
  minor). The G altered scale is
  G--\Aflat--\Bflat--B\nat--C\sharp--\Eflat--F\nat--G, and is sometimes called
  the “diminished whole-tone” scale, since it begins as an octatonic scale and
  ends as a whole-tone scale, or the “super-locrian” scale, the locrian mode
  with a flatted fourth. This sound could be expressed with a number of
  different chord symbols---\h{G7(b9s9s11b13)} or \h{G7(b5s5b9s9)}, for
  example---so jazz musicians typically write “alt.” See
  \headlesscite[70--77]{levine:1995}. We will return to this equivalence
  between chords and scales in Chapter 4.}

\fntext{tfe-2}{%
  The name of the transformation \textsc{ResI} is inspired by Steven Rings's use of
  the transformation ``ResC'' in the first chapter of \headlesscite[25-27]{rings:2011}.}

\fntext{tfe-3}{%
  In fact, we could make similar statements for all of the sets developed in
  this chapter: \Smin would then become ``the set of minor-minor seventh
  chords acting as \sd2, \sd4, and \sd1 in some key'' (loosely speaking; the
  necessary mathematics are easy enough to provide). In most cases, however,
  this level of precision is unnecessary, since the quality of the chord
  uniquely identifies its function.}

\fntext{tfe-4}{%
  For more on diatonic systems generally, see \cite{cloughmyerson:1985}.}

\fntext{tfe-5}{%
  These transformations are all relatively parsimonious, and seem in some way
  related to the \slideS transformation introduced above. We will delay a
  discussion of these parsimonious aspects of these transformations more
  generally until the next chapter.}


\fntext{tfe-6}{%
  ``Ceora'' is perhaps even more diatonic than this space implies, since \h{CM7}
  and \h{BbM7} never appear in the music, while \h{DbM7} does. Thus the
  chord qualities strongly suggest \h{Ab} major: I and IV (\Aflat\ and \Dflat)
  both appear as major sevenths, while ii and iii (\Bflat\ and C) appear only
  as minor seventh chords (unstable ii\tsup{7} chords).}

%}}}

%%% Local Variables: %%%
%%% mode: LaTeX %%%
%%% TeX-master: "diss" %%%
%%% End: %%%
