%% Chapter 3 notes

\fntext{mts-a}{%
  That is, there is no strong voice-leading from \flat{}VII$^7$ to \I in the
  same way as there is in the tritone substitution, \flat{}II$^7$--\I. We will
  return to the voice leading of minor third substitutions below.}

\fntext{mts-1}{%
  Turning Figure \ref{mts:m3-torus} into a torus involves gluing the
  top and bottom edges together and the left and right edges together; the
  dotted line representing perfect fifths then wraps around the surface of the
  torus in a continuous line (as though you had wrapped a barber's pole around a
  doughnut).}

\fntext{mts-2}{%
  In fact, the graphs of the note-based Tonnetz and the m3-torus here are
  isomorphic. This fact introduces some tantalizing possibilities, but none turn
  out to be terribly interesting since, as Richard Cohn has shown, the consonant
  triad is unique among trichords in its capability for parsimonious voice
  leading; \headlesscite[1--7]{cohn:1997}. As such, the m3-torus---made of $(026)$
  trichords---does not show common-tone relations, and the dual graph of the
  m3-torus gives vertices of set class $(0134678\mathrm{T})$.}

\fntext{mts-3}{%
  \cite[246--48]{douthettsteinbach:1998}. The chicken-wire torus is the dual
  graph of the more common note-based Tonnetz; both are shown in
  \cite[Figure 1]{tymoczko:2012}.}

\fntext{mts-4}{%
  The definition of \Smin $\xrightarrow{\mathrm{BD}}$ \Sdom is not given, since
  it is relatively rare in jazz. It models a progression like iv$^7$--\V,
  which is much more common in classical music.}

\fntext{mts-5}{%
  Keith Waters has called this period ``jazz's second practice''
  \parentext{\headlesscite{waters:2013}}. Waters, along with J.~Kent Williams,
  has explored post-tonal jazz harmony using the Tonnetz and hyper-hexatonic
  systems familiar from classical theory; see
  \headlesscite{waterswilliams:2010}.}

\fntext{mts-6}{%
  ``Isotope'' was first recorded on Henderson's album \emph{Inner Urge},
  released in 1965.\nocite{henderson:isotope} It uses a slightly different set
  of changes for the solos than it does for the head (often referred to simply
  as the ``solo changes'' and ``head changes''). This is often the case when
  the head changes are complex, fast-moving, or contain unusual extensions to
  account for specific melody notes. These changes are taken from \emph{The
    Real Book}; the C chord in m.~7 is played as either a major or a dominant
  seventh on the \emph{Inner Urge} recording, so analyzing it as \h{C7} here
  seems reasonable.}

\fntext{mts-7}{%
  We could easily define a BD\tsub{blues} transformation, similar to
  TF\tsub{blues} from the last chapter; for now, we can simply understand this
  variant as BD $\bullet$ \textsc{7th}.}

\fntext{mts-7b}{%
  The question of the meaning of ``function'' is is a difficult one. Here I
  mean the term as Brian Hyer does: ``it is not what a chord \emph{does} that
  matters, but what it \emph{is}: a functional designation names a chord's
  \emph{being}'' \parentext{\headlesscite[109, emphasis original]{hyer:2011}}.}

\fntext{mts-8}{%
  The word ``prolongational'' in this sentence is admittedly problematic,
  given its Schenkerian implications. By using it here I mean only that at
  some deeper level the turnaround is harmonically superfluous, as it occurs
  after the main tonal conclusion of the chorus (an observation confirmed by
  the fact that the turnaround is usually omitted in the last head). I do
  not mean to imply that the turnaround in the last two bars of
  ``Isotope'' is harmonically uninteresting; indeed, it is the most
  distinctive feature of the piece.}

\fntext{mts-9}{%
  Exactly how far apart \h{Bb7} and \h{C7} depends on how one chooses to
  measure voice-leading distance, and whether we consider major-minor sevenths
  in the usual way, as four-note chords, or in the way we have been doing so
  here, as ordered triples of root, third, and seventh. In Jack Douthett's
  Four-Cube Trio, for example, \h{Bb7} and \h{C7} are maximally far apart---4
  semitones; see \cite[157--58]{cohn:2012}.}

\fntext{mts-10}{%
  The notes in these four dominant sevenths form an octatonic collection, and
  have been studied fairly extensively in the literature. See, for example,
  \cite[152--58]{cohn:2012}; \cite[245--46]{douthettsteinbach:1998}; and
  \cite[371]{tymoczko:2011}.}

\fntext{mts-11}{%
  The most obvious differences in mm.~5--12 of the tune are the ``slash chords''
  in mm.~8--9. The chord symbol \h{Em7/A} indicates an E minor seventh chord
  played with an A in the bass; the resulting sound is an \h{A7} chord with a
  suspended fourth (D replaces C\sharp). The older \emph{Real Book} gives the
  same change as \h{A7}sus\h{4}. The only other slight alteration is the addition of
  the ii chord, \h{Dm7}, in m.~10.}


% Major-third spaces
\fntext{maj3-1}{%
  The examples are too numerous to list here, but for an overview, see
  \cite{bribitzerstull:2006}.}

\fntext{maj3-2}{%
  Waters's work in particular has focused extensively on this music, though he
  is hardly alone. See, for example, \cite{julien:2003}; \cite{strunk:2005};
  \cite{waters:2005}; and \cite{waterswilliams:2010}.}

\fntext{maj3-3}{%
  Many people have discussed ``Giant Steps'' in the literature; the most
  substantial work in this area is Matthew Santa's \headlesscite{santa:2003}.
  Guy Capuzzo compares Santa's analysis to one done by Pat Martino in
  \headlesscite{capuzzo:2006}. See also \cite{demsey:1991}; and
  \cite{goodheart:2001}.\nocite{coltrane:giantsteps}}

\fntext{maj3-4}{%
  Frank Samarotto has suggested to me in connection with an unpublished paper
  of his that ``Giant Steps'' is chromatically coherent, while locally diatonic.
  As such, it represents an example of his ``hypothetical'' Type 4 coherence,
  ``in which areas of diatony occur only in local isolation and in which some
  other (presumably post-tonal) coherence might be in effect''
  \parentext{\headlesscite{samarotto:2003}}.}

\fntext{maj3-4a}{%
  These changes are from \cite[30]{levine:piano}. ``Infant Eyes'' appears on
  Shorter's album \emph{Speak No Evil} (1964).\nocite{shorter:evil}}

\fntext{maj3-4b}{%
  Logical arguments could be made for both B and \Eflat as the prevailing key
  of ``Giant Steps.'' Given the organizing influence of the major-third cycle,
  I am not sure the question is so important; the tune uses tonal
  progressions, but may not be \emph{in} a key.}

\fntext{maj3-5}{%
  Exactly how Coltrane devised this substitution set is difficult to say:
  authors have at various times pointed to classical sources---especially
  Nicolas Slonimsky's \emph{Thesaurus of Scales and Melodic Patterns}---as
  well as the music of Thelonious Monk, Dizzy Gillespie, and Tadd Dameron,
  among others. One source that is nearly always cited is the tune ``Have You
  Met Miss Jones?'' (Richard Rodgers/Lorenz Hart), which we analyze in Section
  \ref{subsec:miss-jones}. For a review of these possible origins, see
  \cite[148--57]{demsey:1991}; and \cite[145--47]{porter:1998}.}

\fntext{maj3-6}{%
  Nearly every discussion of Coltrane changes includes the
  ``Countdown''/\,``Tune Up'' pairing. See, for example,
  \cite[159--62]{demsey:1991}; \cite[359--60]{levine:1995}; or many others.
  ``Countdown'' was also recorded on \emph{Giant Steps}, and ``Tune
  Up'' can be heard on \emph{Cookin' with the Miles Davis
  Quintet} (1957).\nocite{davis:cookin}}

\fntext{maj3-7}{%
  Demsey provides a list of third-relations in jazz tunes in an appendix to
  \headlesscite[179--80]{demsey:1991}. Coltrane's famous take on ``Body and Soul'' is
  found on the album \emph{Coltrane's Sound} (1960), which also features two original tunes
  that make prominent use of the major-third cycle: ``Central Park West'' and
  ``Satellite.''\nocite{coltrane:sound}}

\fntext{maj3-8}{%
  % xxx a sentence here on why we're citing Cohn more than anyone else?
  \cite{cohn:2012}. Hexatonic cycles are discussed primarily in
  Cohn's chapter 2, and Weitzmann regions in chapter 4; chapter 5 combines
  these models into a single system which is then used throughout the rest of
  the book. Lewis Porter mentions Weitzmann's treatise on the augmented triad
  as a possible influence on Coltrane \parentext{\headlesscite[146]{porter:1998}}.}

\fntext{maj3-9}{%
  In this figure ``$+$'' indicates major triads and ``$-$'' indicates minor
  triads. ``Cube Dance'' appears in \cite[254]{douthettsteinbach:1998}, and is one of
  Cohn's primary models of triadic space; see \headlesscite[86--109 and
  following]{cohn:2012}.}

\fntext{maj3-10}{%
  This figure is taken from Cohn's book, and has several errors, the most
  important of which is that the C at the 10 o'clock position should be a
  C\sharp, forming a fully-diminished seventh chord. ``4-Cube Trio'' was
  originally devised by Jack Douthett, and is very similar to the ``Power
  Towers'' graphic in \cite[256]{douthettsteinbach:1998} (which omits the
  French sixth chords). For more on its history, see
  \cite[157n15]{cohn:2012}.}

\fntext{maj3-11}{%
  This is a central thesis of Cohn's chapter 7 \parentext{see especially
  \citetitle[148--58]{cohn:2012}}, and figures prominently in Tymoczko's
  geometric theory \parentext{\headlesscite[97--112]{tymoczko:2011}}.}

\fntext{maj3-12}{%
  % This footnote isn't referenced in the text any more, but I want to keep it
  % around here just in case
  My intent here is not to fault these approaches for failing to explain jazz
  harmony, only to demonstrate why they are not sufficient for our purposes
  here. These theories were designed primarily for music of the nineteenth
  century, in which triadic music is the rule and seventh chords the exception.
  In this goal, they generally succeed: many of the unusual harmonic techniques
  in that repertoire \emph{are} well explained by parsimonious voice-leading.
  Jazz, as a music driven by seventh chords, is a different sort of animal,
  and requires a different approach.}

\fntext{maj3-13}{%
  Santa omits the fifth of dominant seventh chords, as we have been doing
  here. The reasons, though, are different: the stated reason is to keep the
  cardinalities of the chords the same, but he does not mention why he chooses
  not to use major seventh chords (rather than triads) and complete dominant
  sevenths, for example. Santa notes later that including the fifth involves
  one of the three missing notes from the nonatonic collection, which is
  acceptable because ``the fourth voice is not essential to the voice leading
  of the cycle'' \parentext{\cite[15]{santa:2003}}.}

\fntext{maj3-14}{%
  The major-third figure represents a kind of cross-section of the
  minor-third torus: to see this clearly, locate the key areas C, \Aflat, and
  E on both Figure \ref{mts:m3-space} and Figure \ref{maj3:maj3-space}.}

\fntext{maj3-15}{%
  In fact, the rest of the figure is unnecessary for ``Giant Steps''; the piece
  is easier to understand using a subgraph of the complete M3-space that
  contains only the \tfo progressions in B, G, and \Eflat.}

\fntext{maj3-16}{%
  These spaces represent all but one of the equal partitions of the octave:
  while we could easily construct a ``whole-tone space,'' it would not have
  very many applications to tonal jazz. In cases where whole-tone
  relationships seem important, it is usually not problematic to consider a
  whole tone as a combination of two perfect fifths, which are readily shown
  in all of the other spaces.}

\fntext{maj3-17}{%
  These changes are again taken from \emph{The Real Book}, and are the
  standard changes for the tune; nearly all recordings agree with this set of
  changes.}

\fntext{maj3-18}{%
  The ``bubble notation'' used in Figure \ref{maj3:miss-jones-bridge} is first
  used in \citetitle[205--6]{lewin:gmit}. It is, as Lewin describes it, a
  ``network-of-networks'': each bubble here represents a single \tfo network
  (the unlabeled arrows are again TF transformations), and these networks are
  connected by larger-scale transpositions. In this figure, the \slideS
  transformations breaks through the bubble itself, and describes a
  transformation directly from \h{Gbmaj7} to \h{Gm7}.}

\fntext{pvl-1}{%
  In addition to including the major seventh chords, this figure corrects some
  errors in Cohn's version (\citetitle[Fig.~7.16, 158]{cohn:2012}): in his
  version, the voice leading to and from the French sixth chords and minor
  seventh chords is incorrect. My thanks to Thomas Cooke-Dickens for helping
  to find many of these errors.}

\fntext{pvl-1b}{%
  The French sixth chord, set class (0268), is of course invariant when
  transposed by tritone, so both roots are given in this figure.}

\fntext{pvl-2}{%
  This is not terribly surprising; our ordered triples are trichords of set
  classes (015), (016), and (026). Richard Cohn has
  shown \parentext{\citetitle{cohn:1997}} that the consonant triad, (037), is
  unique among trichords in its ability to form parsimonious relationships.
  Although his work there does not examine parsimonious relationships among
  members of different set classes, the fact that our three types are not
  nearly even means that we should not expect to find very many of these
  relationships.}

\fntext{pvl-3}{%
  The inverse is needed for \slideS because it was defined as a transformation
  from a major seventh chord to the minor seventh whose root is a half-step
  higher. A cycle that uses the non-inverse \slideS transformation would
  require inverses on both the \textsc{3rd} and \textsc{7th} transformations.}

%%% Local Variables: %%%
%%% mode: latex %%%
%%% TeX-master: "../diss" %%%
%%% End: %%%
